---
title: "Tutorial on Causal Inference Using Machine Learning Methods"
bibliography: book.bib
author: "Kuan Liu"
subtitle: "AI4PH Summer Institute, Aug 18, 2023"
format: 
  html: 
    code-block-bg: true
---

# Welcome

-   Welcome to the causal inference machine learning tutorial!
-   Workshop materials in the github repository [AI4PH2023_CausalWorkshop](https://github.com/Kuan-Liu/AI4PH2023_CausalWorkshop)

# Learning objectives

Causal inference methods, such as the propensity score analysis, have been established to permit causal inference from observational data. In recent years, a growing number of studies have explored the use of machine learning techniques in the causal modelling of complex health data subjected to high-dimensional confounding and complex causal structure. 

The **objective** of this tutorial is to introduce and demonstrate key machine learning methods used in causal inference for cross-sectional data with examples and ready-to-use code in the R programming language. 

By the end of this session, participants should be able to perform causal analysis in R using several machine learning approaches, such as gradient boosting, regression trees, and SuperLearner.

# Tutorial outline

-  Introduction
-  [Section 1: Conventional causal approaches](section1.qmd)
-  [Section 1: Machine learning causal approaches](section2.qmd)
-  Hands-on practice replicating tutorial examples (15-20 mins)


# In preparation for the Tutorial

Participants are required to follow the next steps before the day of the workshop:

1.  Install R and R Studio

    -   Windows operating system
        -   install R, <https://cran.r-project.org/bin/windows/base/>
        -   install RStudio, <https://posit.co/download/rstudio-desktop/#download>
    -   macOS operating system
        -   install R, <https://cran.r-project.org/bin/macosx/>
        -   install RStudio, <https://posit.co/download/rstudio-desktop/#download>

2.  Verify access to the course page, <https://kuan-liu.github.io/AI4PH2023_CausalWorkshop/>

3.  Clone or download the workshop repository: <https://github.com/Kuan-Liu/AI4PH2023_CausalWorkshop>

4.  Install the following R packages
    -  data import and processing and descriptive analysis: tidyverse, tableone, naniar
    -  causal analysis: MatchIt; cobalt; boot; survey; gfoRmula
    -  machine learning: twang, SuperLearner; xgboost; caret; glmnet


# Dataset - The Right Heart Catheterization

For this tutorial, we will be using the same right heart catheterization dataset you have seen this morning. The original JAMA paper [@connors1996effectiveness] and the data csv file can be found in the tutorial repo.

-  We follow Brice's morning session and this tutorial paper [@smith2022introduction], which both used the same rhc dataset, to guide our data processing and causal analysis.


## Data import and processing

```{r echo=TRUE, warning=FALSE, message=FALSE}
library(tidyverse)
data <- read.csv("data/rhc.csv", header=T)

# define exposure variable
data$A <- ifelse(data$swang1 =="No RHC", 0, 1)

# outcome is dth30, a binary outcome measuring survival status at day 30;
data$Y <- ifelse(data$dth30 =="No", 0, 1)
```

## Data visualization on missing values

```{r echo=TRUE, warning=FALSE, message=FALSE, fig.align='center', fig.height=9}
library(naniar)
gg_miss_var(data, facet=A, show_pct = TRUE)
# try changing facet to dth30, this examines missingness by outcome;
```

## Finalizing dataset for causal analysis

```{r echo=TRUE, warning=FALSE, message=FALSE, fig.align='center', fig.height=9}
# we create our analysis data by removing variables with large proportion of missing;
# and variables not used in the analysis;
data2 <- select(data, -c(cat2, adld3p, urin1, swang1,
                         sadmdte, dschdte, dthdte, lstctdte, death, dth30,
                         surv2md1, das2d3pc, t3d30, ptid)) 
data2 <- rename(data2, id = X)

# display data on Quarto page;
library(DT)
data2 %>% datatable(
  rownames = FALSE,
  options = list(
    columnDefs = list(list(className = 'dt-center', 
                      targets = 0:4))))

# verify data structure;
str(data2)
saveRDS(data2,file="data/data2")
```



