---
title: "2. Using machine learning techniques for causal analysis"
bibliography: book.bib
author: "Kuan Liu"
format: 
  html: 
    code-block-bg: true
---

::: callout-tip
## Outlines
-  Propensity score methods machine learning techniques
-  G-computation using machine learning techniques
:::

##  2.1 Proposensity score analysis using machine learning technqiues

###  Super (Machine) Learning

-  Super learning can be used to obtain robust estimator. In a nut-shell it uses loss-function based ML tool and cross-validation to obtain the best prediction of model parameter of interest, based on a **weighted average of a library of machine learning algorithms**.

-  Guide to SuperLearner by Chris Kennedy at https://cran.r-project.org/web/packages/SuperLearner/vignettes/Guide-to-SuperLearner.html

-  List of machine learning algorithms under SuperLearner R package

```{r echo=TRUE, warning=FALSE, message=FALSE}
library(SuperLearner)
listWrappers()
```

###  Using SL with PS approach


```{r echo=TRUE, warning=FALSE, message=FALSE}
require(tidyverse)
data2<-readRDS("data/data2")
covariates <- select(data2, -c(id, A, Y))
baselines <- colnames(covariates)
```
